Illusion Game Validation (Study 2)

### 2022-09-20 study1/study1.Rmd
update study1

---
title: "Illusion Game Validation (Study 1)"
output: html_document
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---

@@

# Preprocessing

@@

We removed `r length(outliers)` participants upon inspection of the average error rage (when close to 50%, suggesting random answers) and/or when the reaction time distribution was implausibly fast.

@@

knitr::kable(dfsub) |> 
  kableExtra::row_spec(which(dfsub$Participant %in% outliers), background  = "#EF9A9A")
```

@@

```{r message=FALSE, warning=FALSE}
dfsub <- df |>
  group_by(Participant) |>
  select(Participant, Age, Sex, Education, Nationality, Ethnicity, Duration, Break_Duration, Screen_Resolution, Screen_Refresh, Device_OS) |>
  slice(1) |>
  ungroup()
```

The final sample included `r report::report_participants(dfsub, age="Age", sex="Sex")`.

@@

@@

@@

```{r message=FALSE, warning=FALSE, cache=cache}
best_models <- function(data) {
  models <- list()
  for(i in 1:1) {
    for(j in 1:1) {
      for(k1 in c("", "_log", "_sqrt", "_cbrt")) { 
        for(k2 in c("")) {  
          for(side in c("", "-side")) {
            name <- paste0("dif", k1, i, "-", "str", k2, j, side)
            # print(name)
            f <- paste0("poly(Illusion_Difference", 
                        k1,
                        ", ",
                        i,
                        ") * poly(Illusion_Strength",
                        k2, 
                        ", ",
                        j, 
                        ") + (1|Participant)")

@@

```{r message=FALSE, warning=FALSE, cache=cache}
cbrt <- function(x) sign(x) * abs(x)**(1/3)

@@

```{r message=FALSE, warning=FALSE, cache=cache}
make_gam <- function(data) {

@@

```{r message=FALSE, warning=FALSE, cache=cache}
best_models(data)
```

@@

```{r message=FALSE, warning=FALSE, cache=cache}
formula <- brms::bf(
  Error ~ cbrt(Illusion_Difference) * Illusion_Strength + 
    (1 | Participant),
  family = "bernoulli"
)

@@

```{r message=FALSE, warning=FALSE, cache=cache}
gam <- make_gam(data)
gam$p
```

@@

```{r message=FALSE, warning=FALSE, cache=cache}
best_models(data)
```

@@

```{r message=FALSE, warning=FALSE, cache=cache}
formula <- brms::bf(
  Error ~ sqrt(Illusion_Difference) * poly(Illusion_Strength, 2) + 
    (1 | Participant),
  family = "bernoulli"
)

@@

```{r message=FALSE, warning=FALSE, cache=cache}
gam <- make_gam(data)
gam$p
```

@@

```{r message=FALSE, warning=FALSE, cache=cache}
best_models(data)
```

@@

```{r message=FALSE, warning=FALSE, cache=cache}
formula <- brms::bf(
  Error ~ sqrt(Illusion_Difference) * Illusion_Strength + 
    (1 | Participant),
  family = "bernoulli"
)

@@

```{r message=FALSE, warning=FALSE, cache=cache}
gam <- make_gam(data)
gam$p
```

@@

```{r message=FALSE, warning=FALSE, cache=cache}
best_models(data)
```

@@

```{r message=FALSE, warning=FALSE, cache=cache}
formula <- brms::bf(
  Error ~ cbrt(Illusion_Difference) * Illusion_Strength + 
    (1 | Participant),
  family = "bernoulli"
)

@@

```{r message=FALSE, warning=FALSE, cache=cache}
best_models(data)
```

@@

```{r message=FALSE, warning=FALSE, cache=cache}
formula <- brms::bf(
  Error ~ log(Illusion_Difference) * poly(Illusion_Strength, 2) + 
    (1 | Participant),
  family = "bernoulli"
)

@@

```{r message=FALSE, warning=FALSE, cache=cache}
gam <- make_gam(data)
gam$p
```

@@

```{r message=FALSE, warning=FALSE, cache=cache}
best_models(data)
```

@@

```{r message=FALSE, warning=FALSE, cache=cache}
formula <- brms::bf(
  Error ~ log(Illusion_Difference) * poly(Illusion_Strength, 2) + 
    (1 | Participant),
  family = "bernoulli"
)

@@

```{r message=FALSE, warning=FALSE, cache=cache}
gam <- make_gam(data)
gam$p
```

@@

```{r message=FALSE, warning=FALSE, cache=cache}
best_models(data)
```

@@

```{r message=FALSE, warning=FALSE, cache=cache}
formula <- brms::bf(
  Error ~ poly(Illusion_Difference, 2) * poly(Illusion_Strength, 2) + 
    (1 | Participant),
  family = "bernoulli"
)

@@

```{r message=FALSE, warning=FALSE, cache=cache}
gam <- make_gam(data)
gam$p
```

@@

```{r message=FALSE, warning=FALSE, cache=cache}
best_models(data)
```

@@

```{r message=FALSE, warning=FALSE, cache=cache}
formula <- brms::bf(
  Error ~ cbrt(Illusion_Difference) * poly(Illusion_Strength, 2) + 
    (1 | Participant),
  family = "bernoulli"
)

@@

```{r message=FALSE, warning=FALSE, cache=cache}
gam <- make_gam(data)
gam$p
```

@@

```{r message=FALSE, warning=FALSE, cache=cache}
best_models(data)
```

@@

```{r message=FALSE, warning=FALSE, cache=cache}
formula <- brms::bf(
  Error ~ Illusion_Difference * poly(Illusion_Strength, 2) + 
    (1 | Participant),
  family = "bernoulli"
)

@@

```{r message=FALSE, warning=FALSE, cache=cache}
gam <- make_gam(data)
gam$p
```

@@

>>>>

---
title: "Illusion Game Validation (Pilot Study)"
author: "Dominique Makowski et al."
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: paper
    code_folding: hide
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
bibliography: "../submission/references.bib"
---

@@

# Introduction

This pilot experiment aimed at gathering some preliminary data to assess if the stimuli generated by *Pyllusion* behaves as expected for each of the 10 illusion types (i.e., whether an increase of task difficulty and illusion strength leads to an increase of errors), and develop an intuition about the magnitude of effects, to refine the stimuli parameters to a more sensible range (i.e., not overly easy and not impossibly hard) for the next study.

In line with open-science standards, all the material (stimuli generation code, experiment code, raw data, analysis script with complementary figures and analyses, preregistration, etc.) is available at [**https://github.com/RealityBending/IllusionGameValidation**](https://github.com/RealityBending/IllusionGameValidation){.uri}.

# Methods

## Procedure

We generated 56 stimuli for each of the 10 illusion types. These stimuli resulted from the combination of 8 linearly-spread levels of task difficulty (e.g., [1, 2, 3, 4, 5, 6, 7], where 1 corresponds to the highest difficulty - i.e., the smallest objective difference between targets) and 7 levels of illusion strength (3 values of strength on the congruent side, 3 on the incongruent side, and 0; e.g., [-3, -2, -1, 0, 1, 2, 3], where negative values correspond to congruent illusion strengths).

The 10 illusion blocks were randomly presented, and the order of the 56 stimuli within the blocks was also randomized. After the first series of 10 blocks, another series was administered (with new randomized orders of blocks and trials). In total, each participant saw 56 different trials per 10 illusion type, repeated 2 times (total = 1120 trials), to which they had to respond "as fast as possible without making errors" (i.e., an explicit double constraint to mitigate the inter-individual variability in the speed-accuracy trade off). The task was implemented using *jsPsych* [@de2015jspsych], and the instructions for each illusion type are available in the experiment code.

@@

We removed `r length(outliers)` participants (in red) upon inspection of the average error rage (when close to 50%, suggesting random answers) and/or when the reaction time distribution was implausibly fast.

knitr::kable(dfsub) |> 
  kableExtra::row_spec(which(dfsub$Participant %in% outliers), background  = "#EF9A9A") |> 
  kableExtra::kable_styling(full_width = TRUE) |> 
  kableExtra::scroll_box(width = "100%", height = "500px")
```

@@

@@

Fifty-two participants were recruited via *Prolific* ([www.prolificacademic.co.uk](www.prolificacademic.co.uk)), a crowd-sourcing platform providing high data quality [@peer2022]. The only inclusion criterion was a fluent proficiency in English to ensure that the task instructions would be well-understood. Participants were incentivised with a reward of about \textsterling 7.5 for completing the task, which took about 50 minutes to finish. Demographic variables (age, gender, and ethnicity) were self-reported on a voluntary basis.

We removed 6 participants upon inspection of the average error rate (when close to 50%, suggesting random answers), and when the reaction time distribution was implausibly fast. For the remaining participants, we discarded blocks where the error rate was higher than 50% (possibly indicating that instructions got misunderstood; e.g., participants were selecting the shorter line instead of the longer one). Finally, we removed 692 (1.37%) trials based on an implausibly short or long response time (\< 150 ms or \> 3000 ms).

The final sample included 46 participants (Mean age = 26.7, SD = 7.7, range: [19, 60]; Sex: 39.1% females, 56.5% males, and 4.4% other).

```{r message=FALSE, warning=FALSE, results='hide'}
dfsub <- df |>
  group_by(Participant) |>
  select(Participant, Age, Sex, Education, Nationality, Ethnicity, Duration, Break_Duration, Screen_Resolution, Screen_Refresh, Device_OS) |>
  slice(1) |>
  ungroup()

report::report_participants(dfsub, age="Age", sex="Sex")
```

@@

## Data Analysis

The analysis focused on the probability of errors as the main outcome variable. For each illusion, we started by visualizing the average effect of task difficulty and illusion strength to gain some intuition on the underlying generative model. Next, we tested the performance of various logistic models differing in their specifications, such as: with or without a transformation of the task difficulty (log, square root or cubic root), with or without a 2nd order polynomial term for the illusion strength, and with or without the illusion side (up *vs.* down or left *vs.* right) as an additional predictor. We then fitted the best performing model under a Bayesian framework, and compared its visualization with that of a General Additive Model (GAM), which has an increased ability of mapping underlying potential non-linear relationships (at the expense of model simplicity).

The analysis was carried out using *R 4.2* [@RCoreTeam2022], *brms* [@Burkner2017], the *tidyverse* [@wickham2019], and the *easystats* collection of packages [@bayestestRArticle; @correlationArticle; @performanceArticle; @insightArticle].

@@

## Summary

The statistical models suggested that the effect of task difficulty had a cubic relationship with error rate for the Delboeuf and Ebbinghaus illusions (both composed of circular shapes), square relationship for the Rod and Frame and Vertical-Horizontal illusions, cubic relationship for the Zöllner and Poggendorff illusions, exponential relationship for the White illusion, cubic relationship for the Müller-Lyer and Ponzo illusions (both based on line lengths), and linear relationship for the Contrast illusion. All models suggested a significant effect of illusion strength and task difficulty. See details and figures in the analysis script.

@@

```{r delboeuf_bestmodels, message=FALSE, warning=FALSE, cache=cache}
best_models <- function(data) {
  models <- list()
  for(i in 1:1) {
    for(j in 1:1) {
      for(k1 in c("", "_log", "_sqrt", "_cbrt")) { 
        for(k2 in c("")) {  
          for(side in c("", "-side")) {
            name <- paste0("dif", k1, i, "-", "str", k2, j, side)
            # print(name)
            f <- paste0("poly(Illusion_Difference", 
                        k1,
                        ", ",
                        i,
                        ") * poly(Illusion_Strength",
                        k2, 
                        ", ",
                        j, 
                        ") + (1|Participant)")

@@

```{r delboeuf_model, message=FALSE, warning=FALSE, cache=cache}
cbrt <- function(x) sign(x) * abs(x)**(1/3)

@@

```{r delboeuf_gam, message=FALSE, warning=FALSE, cache=cache}
make_gam <- function(data) {

@@

```{r ebbinghaus_bestmodels, message=FALSE, warning=FALSE, cache=cache}
best_models(data)
```

@@

```{r ebbinghausmodel, message=FALSE, warning=FALSE, cache=cache}
formula <- brms::bf(
  Error ~ cbrt(Illusion_Difference) * Illusion_Strength + 
    (1 | Participant),
  family = "bernoulli"
)

@@

```{r ebbinghaus_gam, message=FALSE, warning=FALSE, cache=cache}
gam <- make_gam(data)
gam$p
```

@@

```{r rodframe_bestmodels, message=FALSE, warning=FALSE, cache=cache}
best_models(data)
```

@@

```{r rodframe_model, message=FALSE, warning=FALSE, cache=cache}
formula <- brms::bf(
  Error ~ sqrt(Illusion_Difference) * poly(Illusion_Strength, 2) + 
    (1 | Participant),
  family = "bernoulli"
)

@@

```{r rodframe_gam, message=FALSE, warning=FALSE, cache=cache}
gam <- make_gam(data)
gam$p
```

@@

```{r verticalhorizontal_bestmodels, message=FALSE, warning=FALSE, cache=cache}
best_models(data)
```

@@

```{r verticalhorizontal_model, message=FALSE, warning=FALSE, cache=cache}
formula <- brms::bf(
  Error ~ sqrt(Illusion_Difference) * Illusion_Strength + 
    (1 | Participant),
  family = "bernoulli"
)

@@

```{r verticalhorizontal_gam, message=FALSE, warning=FALSE, cache=cache}
gam <- make_gam(data)
gam$p
```

@@

```{r zollner_bestmodels, message=FALSE, warning=FALSE, cache=cache}
best_models(data)
```

@@

```{r zollner_model, message=FALSE, warning=FALSE, cache=cache}
formula <- brms::bf(
  Error ~ cbrt(Illusion_Difference) * Illusion_Strength + 
    (1 | Participant),
  family = "bernoulli"
)

@@

```{r white_bestmodels, message=FALSE, warning=FALSE, cache=cache}
best_models(data)
```

@@

```{r white_model, message=FALSE, warning=FALSE, cache=cache}
formula <- brms::bf(
  Error ~ log(Illusion_Difference) * poly(Illusion_Strength, 2) + 
    (1 | Participant),
  family = "bernoulli"
)

@@

```{r white_gam, message=FALSE, warning=FALSE, cache=cache}
gam <- make_gam(data)
gam$p
```

@@

```{r mullerlyer_bestmodels, message=FALSE, warning=FALSE, cache=cache}
best_models(data)
```

@@

```{r mullerlyer_model, message=FALSE, warning=FALSE, cache=cache}
formula <- brms::bf(
  Error ~ log(Illusion_Difference) * poly(Illusion_Strength, 2) + 
    (1 | Participant),
  family = "bernoulli"
)

@@

```{r mullerlyer_gam, message=FALSE, warning=FALSE, cache=cache}
gam <- make_gam(data)
gam$p
```

@@

```{r ponzo_bestmodels, message=FALSE, warning=FALSE, cache=cache}
best_models(data)
```

@@

```{r ponzo_model, message=FALSE, warning=FALSE, cache=cache}
formula <- brms::bf(
  Error ~ poly(Illusion_Difference, 2) * poly(Illusion_Strength, 2) + 
    (1 | Participant),
  family = "bernoulli"
)

@@

```{r ponzo_gam, message=FALSE, warning=FALSE, cache=cache}
gam <- make_gam(data)
gam$p
```

@@

```{r poggendorff_bestmodels, message=FALSE, warning=FALSE, cache=cache}
best_models(data)
```

@@

```{r poggendorff_model, message=FALSE, warning=FALSE, cache=cache}
formula <- brms::bf(
  Error ~ cbrt(Illusion_Difference) * poly(Illusion_Strength, 2) + 
    (1 | Participant),
  family = "bernoulli"
)

@@

```{r poggendorff_gam, message=FALSE, warning=FALSE, cache=cache}
gam <- make_gam(data)
gam$p
```

@@

```{r contrast_bestmodels, message=FALSE, warning=FALSE, cache=cache}
best_models(data)
```

@@

```{r contrast_model, message=FALSE, warning=FALSE, cache=cache}
formula <- brms::bf(
  Error ~ Illusion_Difference * poly(Illusion_Strength, 2) + 
    (1 | Participant),
  family = "bernoulli"
)

@@

```{r contrast_gam, message=FALSE, warning=FALSE, cache=cache}
gam <- make_gam(data)
gam$p
```

@@

# Discussion

This pilot study provided a clearer understanding of the magnitude of the parametric effects at stake and the type of interaction between them. Furthermore, it allowed us to better understand and test the stimuli generated by *Pyllusion*, as well as uncover incidental bugs and technical issues (for instance, the specification direction of the illusion strength was reversed for a few illusions), which were fixed in a new software release. Crucially, this study allowed us to refine the range of task difficulty and illusion strength values in order to maximize information gain.

In most illusions, the task difficulty exhibited monotonic power-law scaled effects, which is in line with the psychophysics literature on perceptual decisions [@ditzinger2010; @shekhar2021; @bogacz2006]. One notable result was the illusion effect pattern for the Zöllner illusion, which suggested a non-linear relationship. By generating a wider range of illusion strength values, the [**next study**](https://github.com/RealityBending/IllusionGameValidation) will attempt at clarifying this point.

# References

### 2022-09-20 study2/study2.Rmd
update study1

```{r message=FALSE, warning=FALSE}
data.frame(Participant = c("Total"), t(sapply(dfsub[2:ncol(dfsub)], mean, na.rm=TRUE))) |> 
  rbind(dfsub) |> 
  knitr::kable() |>
  kableExtra::row_spec(1, italic = TRUE) |> 
  kableExtra::row_spec(which(dfsub$Participant %in% outliers) + 1, background = "#EF9A9A") |> 
  kableExtra::row_spec(which(dfsub$Participant %in% partial_outliers) + 1, background = "#FFECB3")
```

>>>>

```{r message=FALSE, warning=FALSE}
data.frame(Participant = c("Total"), t(sapply(dfsub[2:ncol(dfsub)], mean, na.rm=TRUE))) |> 
  rbind(dfsub) |> 
  knitr::kable() |>
  kableExtra::row_spec(1, italic = TRUE) |> 
  kableExtra::row_spec(which(dfsub$Participant %in% outliers) + 1, background = "#EF9A9A") |> 
  kableExtra::row_spec(which(dfsub$Participant %in% partial_outliers) + 1, background = "#FFECB3") |> 
  kableExtra::kable_styling(full_width = TRUE) |> 
  kableExtra::scroll_box(width = "100%", height = "500px")
```
